`r opts_chunk$set(cache=TRUE)`

# Winter Olympics Medal Example

## Import Dataset
```{r import_data, warning=FALSE}
library(xtable)
options(stringsAsFactors=FALSE)
medals <- read.csv("data/medals.csv")
medals$Year <- as.numeric(medals$Year)
medals <- medals[!is.na(medals$Year), ]       
```

The Olympic Medals data frame includes `r nrow(medals)` medals from `r min(medals[["Year"]])` to  `r max(medals[["Year"]])`.  The data was sourced from [The Guardian Data Blog](http://www.guardian.co.uk/news/datablog/2010/feb/11/winter-olympics-medals-by-country).


## Total Medals by Year
```{r prepare_total_medals}
# http://www.math.mcmaster.ca/~bolker/emdbook/chap3A.pdf
x <- aggregate(medals$Year, list(Year = medals$Year), length)
names(x) <- c("year", "medals") 
x$pos <- seq(x$year) 
fit <- nls(medals ~ a * pos ^ b + c, x, start = list(a=10, b=1, c = 50))
```

In general over the years the number of Winter Olympic medals awarded has increased.  In order to model this relationship, year was converted to ordinal position.  A three parameter power function seemed plausible, $latex y = ax^b + c$, where $latex y$ is total medals awarded and $latex x$ is the ordinal position of the olympics starting at one.  The best fitting parameters by least-squares were

$$latex 
`r I(round(coef(fit)["a"], 3))` 
x^{`r I(round(coef(fit)["b"], 3))` + `r I(round(coef(fit)["c"], 3))`}.
$$

The figure displays the data and the line of best fit for the model. The model predicts that 2010, 2014, and 2018 would have  `r round(predict(fit, data.frame(pos=max(x[["pos"]]) + 1)))`,  `r round(predict(fit, data.frame(pos=max(x[["pos"]]) + 2)))`, and `r round(predict(fit, data.frame(pos=max(x[["pos"]]) + 3)))` medals  respectively.
                      
```{r figure_of_medals, fig=TRUE, tidy=FALSE}
plot(medals ~ pos, x,  las = 1, 
		ylab = "Total Medals Awarded", 
		xlab = "Ordinal Position of Olympics",
        main="Total medals awarded 
     by ordinal position of Olympics with
     predicted three parameter power function fit displayed.",
		las = 1,
		bty="l")
lines(x$pos, predict(fit))
```
        

# Gender Ratio by Year
```{r fgenderRatioByYear_setup_data}
medalsByYearByGender <- aggregate(medals$Year, 
    list(Year = medals$Year, Event.gender = medals$Event.gender), length)
medalsByYearByGender <- medalsByYearByGender[medalsByYearByGender$Event.gender 
    != "X", ]
propf <- list()
propf$prop <- medalsByYearByGender[medalsByYearByGender$Event.gender == "W", "x"] / (
			medalsByYearByGender[medalsByYearByGender$Event.gender == "W", "x"] +
			medalsByYearByGender[medalsByYearByGender$Event.gender == "M", "x"])
propf$year <- medalsByYearByGender[medalsByYearByGender$Event.gender == "W", "Year"]
propf$propF <- format(round(propf$prop, 2))

propf$table <- with(propf, 	cbind(year, propF))
colnames(propf$table) <- c("Year", "Prop. Female")
```

The figure shows the number of medals won by males and  females by year. The table shows the proportion of medals awarded to females by year. It shows a generally similar pattern for males and females. Medals increase gradually until around the late 1980s after which the rate of increase accelerates. However, females started from a much smaller base. Thus, both the absolute difference and the percentage difference has decreased over time to the point where in 2006 `r as.numeric(propf[["propF"]][length(propf[["propF"]])])* 100` of medals were won by females.

```{r fgenderRatioByYear_figure, fig=TRUE}
plot(x ~ Year, medalsByYearByGender[medalsByYearByGender$Event.gender == "M", ], 
    ylim = c(0,max(x)), pch = "m", col = "blue", las = 1,
    ylab = "Total Medals Awarded", bty="l",
     main="Total Medals Won by Gender and Year")
points(medalsByYearByGender[medalsByYearByGender$Event.gender == "W", "Year"],
    medalsByYearByGender[medalsByYearByGender$Event.gender == "W", "x"],
    col = "red", pch = "f")
```




```{r table_prop_female, results='asis', tidy=FALSE}
print(xtable(propf$table,
             caption="Proportion of Medals that were awarded to Females by Year"), 
      type="html", 
      caption.placement="top",
      html.table.attributes='align="center"')
```   

# Countries with the Most Medals
```{r country_most_medals}
cmm <- list()
cmm$medals <- sort(table(medals$NOC), dec = TRUE)
cmm$country <- names(cmm$medals)
cmm$prop <- cmm$medals / sum(cmm$medals)
cmm$propF <- paste(round(cmm$prop * 100, 2), "%", sep ="")

cmm$row1 <- c("Rank", "Country", "Total", "%")
cmm$rank <- seq(cmm$medals)
cmm$include <- 1:10
                               
cmm$table <- with(cmm,
		rbind(cbind(rank[include], country[include],
						medals[include], propF[include])))
colnames(cmm$table) <- cmm$row1
```

Norway has won the most medals with `r cmm[["medals"]][1]` (`r round(cmm[["prop"]][1]* 100, 2)`%). The table shows the top 10.  Russia, USSR, and EUN (Unified Team in 1992 Olympics) have a combined total of `r sum(cmm$medals[c("RUS", "URS", "EUN")])`.  Germany, GDR, and FRG have a combined medal total of  `r sum(cmm$medals[c("GER", "GDR", "FRG")])`.



```{r table_rankings_display, results='asis', tidy=FALSE}
print(xtable(cmm$table, caption="Rankings of Medals Won by Country"), 
      "html", include.rownames=FALSE, caption.placement='top',
      html.table.attributes='align="center"')
```

# Proportion of Gold Medals by Country
Looking only at countries that have won more than 50 medals in the dataset, the figure shows that the proportion of medals won that were gold, silver, or bronze.

```{r proportion_gold, fig=TRUE, tidy=FALSE}
NOC50Plus <- names(table(medals$NOC)[table(medals$NOC) > 50])
medalsSubset <- medals[medals$NOC %in% NOC50Plus, ]
medalsByMedalByNOC <- prop.table(table(medalsSubset$NOC, medalsSubset$Medal), 
                                 margin = 1)
medalsByMedalByNOC <- medalsByMedalByNOC[order(medalsByMedalByNOC[, "Gold"], 
         decreasing = TRUE), c("Gold", "Silver", "Bronze")]
barplot(round(t(medalsByMedalByNOC), 2), horiz = TRUE, las = 1, 
		col=c("gold", "grey71", "chocolate4"), 
		xlab = "Proportion of Medals",
        main="Proportion of medals won that were gold, silver or bronze.")
```



# How many different countries have won medals by year?
```{r number_of_countries_prepare}
listOfYears <- unique(medals$Year)                   
names(listOfYears) <- unique(medals$Year)
totalNocByYear <- sapply(listOfYears,  function(X) 
      length(table(medals[medals$Year == X, "NOC"])))
```

The figure shows the total number of countries winning medals by year.

```{r figure_total_medals, fig=TRUE}
plot(x= names(totalNocByYear), totalNocByYear, 
    ylim = c(0, max(totalNocByYear)),
		las = 1,
    xlab = "Year",
    main = "Total Number of Countries Winning Medals By Year",
    ylab = "Total Number of Countries",
    bty = "l")
```


# Australia at the Winter Olympics
```{r ausmedals}
ausmedals <- list()
ausmedals$data <- medals[medals$NOC == "AUS", ]
ausmedals$data <- ausmedals$data[, c("Year", "City", "Discipline", "Event",
"Medal")]
ausmedals$table <- ausmedals$data
```

Given that I am an Australian I decided to have a look at the Australian medal count. Australia does not get a lot of snow.  Up to and including 2006, Australia has won  `r nrow(ausmedals$data)` medals. It won its first medal in  `r min(ausmedals$data$Year)`. Of the `r nrow(ausmedals$data)` medals,  `r sum(ausmedals$data$Medal == "Bronze")` were bronze, `r sum(ausmedals$data$Medal == "Silver")` were silver, and `r sum(ausmedals$data$Medal == "Gold")` were gold.  The table lists each of these medals.

```{r table_australian_medal_list, results='asis', tidy=FALSE}
print(xtable(ausmedals$table, 
             caption='List of Australian Medals',
             digits=0),
      type='html', 
      caption.placement='top', 
      include.rownames=FALSE,
      html.table.attributes='align="center"') 
```   

# Ice Hockey
```{r icehockey_prepare}
icehockey <- medals[medals$Sport == "Ice Hockey" & 
    medals$Event.gender == "M" &
    medals$Medal == "Gold",  ]
icehockeyf <- medals[medals$Sport == "Ice Hockey" & 
    medals$Event.gender == "W" &
    medals$Medal == "Gold",  ]

# names(table(icehockey$NOC)[table(icehockey$NOC) > 1])
```

The following are some statistics about Winter Olympics Ice Hockey up to and including the 2006 Winter Olympics.  

* Out of the  `r length(unique(medals$Year))`  Winter Olympics that have been staged,  Mens Ice Hockey has been held in  `r nrow(icehockey)` and the Womens in `r nrow(icehockeyf)`.
* The USSR has won the most mens gold medals with `r sum(icehockey$NOC == "URS")` golds. It goes up to `r sum(icehockey$NOC %in% c("EUN", "URS"))` if the 1992 Unified Team is included. 
* Canada has the second most golds with  `r sum(icehockey$NOC == "CAN")`. 
* After that the only two nations to win more than one gold are  Sweden (`r sum(icehockey$NOC == "SWE")` golds) and the United States (`r sum(icehockey$NOC == "USA")` golds).
*  The  table shows the countries who won gold and silver medals by year.
* In the case of the Women's Ice Hockey, Canada has won `r table(icehockeyf$NOC)["CAN"]` and the United States has won `r table(icehockeyf$NOC)["USA"]`.

```{r icehockey_gold_silver, results='asis', tidy=FALSE}
icehockeygs <- medals[medals$Sport == "Ice Hockey" & 
    medals$Event.gender == "M" &
    medals$Medal %in% c("Silver", "Gold"),  c("Year", "Medal", "NOC")]
icetab <- list()
icetab$data <- reshape(icehockeygs, idvar="Year", timevar="Medal",
    direction="wide")
names(icetab$data) <- c("Year", "Gold", "Silver")

print(xtable(icetab$data, 
             caption ="Country Winning Gold and Silver Medals by Year in Mens Ice Hockey", 
             digits=0), 
      type="html",     
      include.rownames=FALSE,
      caption.placement="top",
      html.table.attributes='align="center"')

```

# Reflections on the Conversion Process
* Markdown versus LaTeX: 
     * I prefer performing analyses with Markdown than I do with LateX. 
     * Markdown is easier to type than LaTeX. 
     * Markdown is easier to read than LaTeX.
     * It is easier with Markdown to get started with analyses.
     * Many analyses are only presented on the screen and as such page breaks in LaTeX are a nuisance. This extends to many features of LaTeX such as headers, figure and table placement, margins, table formatting, partiuclarly for long or wide tables, and so on.
     * That said, journal articles, books, and other artefacts that are bound to the model of a printed page are not going anywhere. 
     * Furthermore, bibliographies, cross-references, elaborate control of table appearance, and more are all features which LaTeX makes easier than Markdown.
* R Markdown to Sweave LaTeX:
    * The more common conversion task that I can imagine is taking some simple analyses in R Markdown and having to convert them into knitr LaTeX in order to include the content in a journal article.
    * The first time I converted between the formats, it was good to do it in a relatively manual way to get a sense of all the required changes; however, if I had a large document or was doing the task on subsequent occasions, I would look at more automated solutions using string replacement tools (e.g., sed, or even just replacement commands in a text editor such as Vim), and markup conversion tools (e.g., pandoc).
    * Perhaps if the formats get popular enough, developers will start to build dedicated conversion tools.


# Additional Resources
If you liked this post, you may want to subscribe to the [RSS feed of my blog](http://feeds.feedburner.com/jeromyanglim). Also see:
* This post on [Getting Started with R Markdown, knitr, and Rstudio 0.96](http://jeromyanglim.blogspot.com/2012/05/getting-started-with-r-markdown-knitr.html)
* This post for another [Example Reproducible Report using R Markdown which analyses California Schools Test Data](http://jeromyanglim.blogspot.com/2012/05/example-reproducile-report-using-r.html)
* These [Assorted posts using Sweave](http://jeromyanglim.blogspot.com.au/search/label/Sweave)
* The [knitr](http://yihui.name/knitr/) home page and [knitr options page](http://yihui.name/knitr/options).
* the [xtable LaTeX table gallery](http://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf) which can also be used to generate HTML tables for inclusion in Markdown.