
## Prepare for analyses
```{r setup}
set.seed(1234)
library(ggplot2)
library(lattice)
```


## Basic console output
To insert an R code chunk, you can type it manually or just press `Chunks - Insert chunks` or use the shortcut key. This will produce the following code chunk:

    ```{r}
    
    ```

Pressing tab when inside the braces will bring up code chunk options.

The following R code chunk labelled `basicconsole` is as follows:

    ```{r basicconsole}
    x <- 1:10
    y <- round(rnorm(10, x, 1), 2)
    df <- data.frame(x, y)
    df
    ```
The code chunk input and output is then displayed as follows:

```{r basicconsole}
x <- 1:10
y <- round(rnorm(10, x, 1), 2)
df <- data.frame(x, y)
df
```

## Plots
Images generated by `knitr` are saved in a figures folder. However, they also appear to be represented in the HTML output using a [data URI scheme]( http://en.wikipedia.org/wiki/Data_URI_scheme). This means that you can paste the HTML into a blog post or discussion forum and you don't have to worry about finding a place to store the images; they're embedded in the HTML.

### Simple plot
Here is a basic plot using base graphics:

    ```{r simpleplot}
    plot(x)
    ```

```{r simpleplot}
plot(x)
```

Note that unlike traditional Sweave, there is no need to write `fig=TRUE`.


### Multiple plots
Also, unlike traditional Sweave, you can include multiple plots in one code chunk:

    ```{r multipleplots}
    boxplot(1:10~rep(1:2,5))
    plot(x, y)
    ```

```{r multipleplots}
boxplot(1:10~rep(1:2,5))
plot(x, y)
```

### `ggplot2` plot
Ggplot2 plots work well:

```{r ggplot2ex}
qplot(x, y, data=df)
```

### `lattice` plot
As do lattice plots:

```{r latticeex}
xyplot(y~x)
```

Note that unlike traditional Sweave, there is no need to print lattice plots directly.

## R Code chunk features
### Create Markdown code from R
The following code hides the command input (i.e., `echo=FALSE`), and outputs the content directly as code (i.e., `results=asis`, which is similar to `results=tex` in Sweave).


    ```{r dotpointprint, results='asis', echo=FALSE}
    cat("Here are some dot points\n\n")
    cat(paste("* The value of y[", 1:3, "] is ", y[1:3], sep="", collapse="\n"))
    ```

```{r dotpointprint, results='asis', echo=FALSE}
cat("Here are some dot points\n\n")
cat(paste("* The value of y[", 1:3, "] is ", y[1:3], sep="", collapse="\n"))
```

### Create Markdown table code from R
    ```{r createtable, results='asis', echo=FALSE}
    cat("x | y", "--- | ---", sep="\n")
    cat(apply(df, 1, function(X) paste(X, collapse=" | ")), sep = "\n")
    ```

```{r createtable, results='asis', echo=FALSE}
cat("x | y", "--- | ---", sep="\n")
cat(apply(df, 1, function(X) paste(X, collapse=" | ")), sep = "\n")
```

### Control output display
The folllowing code supresses display of R input commands (i.e., `echo=FALSE`)
and removes any preceding text from console output (`comment=""`; the default is `comment="##"`).

    ```{r echo=FALSE, comment="", echo=FALSE}
    head(df)
    ```

```{r echo=FALSE, comment="", echo=FALSE}
head(df)
```



### Control figure size
The following is an example of a smaller figure using `fig.width` and `fig.height` options.

    ```{r smallplot, fig.width=3, fig.height=3}
    plot(x)
    ```

```{r smallplot, fig.width=3, fig.height=3}
plot(x)
```